Informing policy via dynamic models: Cholera in Haiti

### 2024-02-09 r1/response.tex

A hyphen has been added for all instances of the phrase: ``model-based".

\report{
  17. Lines 576-577. ``We notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (S5 Text)". After this fragment, please summarise in one or two sentences what it will be found in S5 Text.
}

\report{
  18. Lines 599-601. ``The decision not to do this partially explains the unsuccessful forecasts of Lee et al. [1]: their Table S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera". The fragment before the colon says that Lee was unsuccessful. However, the fragment after the colon says that was in part successful. Please clarify.
}

In their article, Lee et al.~\cite{lee20} only presented forecasts that were based on simulations from initial conditions. These forecasts were generally inconsistent with the absence of cholera cases from 2019-2022, and do not account for recent evidence of cholera dynamics. In their supplement material, however, they provide a table (S7) in which they show that if they condition the simulations on most recent data (0 cases from 2019-2020), their forecasts predict cholera elimination. This second effort that was only presented in the supplement material is more consistent with what we suggest is correct practice: forecasts of a system should start using latent states that are consistent with available data. We have expanded this argument in order to clarify this point:

\article{\editForecastOne}

\article{\editForecastTwo}

@@

\report{
  19. Line 631. Since Model 1 accounts for infections at the national level, how are scenarios V1 and V2 handled?
}

We followed the same approach as Lee et al.~\cite{lee20}, which was requiring that the same number of vaccinations are administered at the national level as they would have been in the spatially-explicit vaccination campaigns. 
The following sentence has been added to clarify this:

\article{\editModVacc}

### 2024-02-09 edits.tex
Changed discussion section to use edits, also made edits and response to reviewers comments for clarity

\newcommand\editModelingGuidanceSecond{Based on our assessment and literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling. Concretly, \cite{lee20} follows at least four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}. Determining the level of adherence to the first principle, \emph{stakeholder engagement}, is difficult based solely on the article. Despite this, the inconsistency between their forecasts and the cholera incidence from 2019 to 2022 suggests that existing recommendations and standards related to policy-driven modeling may be insufficient. \cite{saltelli19} suggests that improvements in model model-based outputs may be obtained by developing structures and standards based on statistical principles. Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework. 
}

@@

>>>>

\newcommand\editModelingGuidanceSecond{Based on our assessment and literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling. Concretly, \cite{lee20} follows at least four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}. Determining the level of adherence to the first principle, \emph{stakeholder engagement}, is difficult based solely on the article. Despite this, the inconsistency between their forecasts and the cholera incidence from 2019 to 2022 suggests that existing recommendations and standards related to policy-driven modeling may be insufficient. \cite{saltelli19} suggests that improvements in model model-based outputs may be obtained by developing structures and standards based on statistical principles. Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework.}

@@

\newcommand\editDiscussionOne{In our cholera case study, we highlight the importance of model diagnostics; they are instrumental in enhancing model fits, thus leading to more trustworthy conclusions. We compared the Models 1--3 of \cite{lee20} to simple statistical benchmarks, enabling us to identify deficiencies in the models that could be improved. In the case of Model~3, comparing a fitted model to a benchmark and then following up by questioning why the model did not describe the data as efficiently as its benchmark helped us discern that the model fell short in explaining the surge in cholera transmission following Hurricane Matthew. This led us to suggest a model modification. Because Model~1 outperformed its benchmark model, we were able to confidently scrutinize the implications of the fitted model. We found that the seasonality term of Model~1 closely mirrors the seasonal rainfall patterns in Haiti, reaffirming alternative evidence of the importance of rainfall as a driver of cholera infection and bolstering our confidence in Model~1's ability to capture the critical dynamics of the cholera outbreak.}

\newcommand\editDiscussionTwo{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric like AIC. Nested model variations are particularly useful as they enable formal statistical testing of the nested features. The cholera case study demonstrated the advantage of such a model variation in all three models. For instance, we tested the inclusion of a trend in transmission rate over time in Model~1; in Model~2, we considered adding an additional phase parameter, which enables a shift in the seasonal peaks of cholera infections; in Model~3, we included additional parameters such as those associated with Hurricane Matthew. In all these cases, we found strong statistical evidence in favor of these model variations over their simpler counterparts.}

\newcommand\editDiscussionThree{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model. Mathematical models necessarily simplify complex dynamical systems in order to make inference possible. Decision-making about which features of a system need to be incorporated in a model and which should be excluded can significantly influence the model-based conclusions. However, these decisions can be tested using a likelihood-based framework. Standard model decisions cover deterministic versus stochastic modeling and the decision to include or exclude a meta-population structure. Model benchmarks can assist in determining the level of complexity to be used in the model, enabling researchers to gauge a model's relative performance against a better-understood statistical model.}

\newcommand\editDiscussionFour{The three models in this case study offer a unique combination of decisions of model complexity. Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences. In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm. One of only a few examples of fitting a model with this level of complexity, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}

\newcommand\editDiscussionFive{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance. This is particularly true for long time series, as the latent states from simulations from initial conditions may be more likely to diverge from the truth as time increases. One way to address this issue is by simulating forward from the filtering distribution at the last available time point, as this proceedure considers available evidence about the latent states of a system under the framework of the dynamic model. This type of forecasting, however, is not possible using a deterministic model. Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}. Despite these limitations, forecasts from deterministic models can provide valuable insights on potential future trends, though researchers should be transparent about these limitations when presenting forecasts from deterministic models.}

\newcommand\editModVacc{Because Model~1 only accounts for national level disease dynamics, the pre-determined department-specific vaccination campaigns are carried out by assuming the vaccines are administered in one week to the same number of individuals that would have obtained vaccines if explicitly administered to the specific departments. We refer readers to \cite{lee20} and the accompanying supplement material for more details.}

\newcommand\editForecastOne{Forecasts of a dynamic system should ideally be based on the most recent information available as it is likely to be more relevant than older data. While this assertion may seem self-evident, it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance. Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}. Lee et al.~\cite{lee20} chose to obtain forecasts from all of their models by simulating forward from initial conditions.
This decision is possibly as a result of using a deterministic model: forecasts from different models may only be considered comparable only if they are obtained in the same way, which must be by simulating from initial conditions because Model~2 is deterministic.}

\newcommand\editForecastTwo{In contrast, for non-deterministic Models~1 and 3, we obtain forecasts by simulating future values using latent states that are harmonious with the most recent data. This is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$. The decision to obtain model forecasts from initial conditions partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}. Table~S7 in their supplement material, which contains results that were not discussed in their main article, shows that the subset of their simulations which were consistent with observing zero cases from 2019-2020 were also more consistent with the disappearance of cholera from Haiti from 2019-2022. These results support our argument: forecasts of a dynamic system should start with latent states that are consistent with available data.}

### 2024-02-09 ms.Rnw
Changed discussion section to use edits, also made edits and response to reviewers comments for clarity

\TODO{Re-phrase this paragraph, as suggested by one of the reviewers.}
Recent information about a dynamic system should be more relevant for a forecast than older information.
This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
\new{For Models~1 and 3---which are non-deterministic---}we obtain forecasts by simulating future values using latent states that are consistent with the most recent observations; this is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.

@@

Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
The results of these projections are summarized in Figs. \ref{fig:elimProbs}.

@@

\new{In our cholera case study, we highlight the importance of model diagnostics; they are instrumental in enhancing model fits, thus leading to more trustworthy conclusions.
We compared the Models 1--3 of \cite{lee20} to simple statistical benchmarks, enabling us to identify deficiencies in the models that could be improved.
In the case of Model~3, comparing a fitted model to a benchmark and then following up by questioning why the model did not describe the data as efficiently as its benchmark helped us discern that the model fell short in explaining the surge in cholera transmission following Hurricane Matthew.
This led us to suggest a model modification.
Because Model~1 outperformed its benchmark model, we were able to confidently scrutinize the implications of the fitted model.
We found that the seasonality term of Model~1 closely mirrors the seasonal rainfall patterns in Haiti, reaffirming alternative evidence of the importance of rainfall as a driver of cholera infection and bolstering our confidence in Model~1's ability to capture the critical dynamics of the cholera outbreak. }

% \new{In any study involving dynamic models, it is important to consider, test, and present more than one model framework or variant.
% \eic{I'M NOT SURE THIS IS A CONCLUSION. LEE ET AL STUDIED MULTIPLE MODELS BECAUSE THAT WAS THE GOAL OF THAT PAPER. MOST OTHER PAPERS STUDY ONE MODEL. WE WOULD LIKE RESULTS FROM DIFFERENT PAPERS TO HAVE OBJECTIVE MEASURES OF FIT THAT CAN BE COMPARED BETWEEN PAPERS, BUT I DON'T THINK WE WANT TO SAY THAT ALL PAPERS HAVE TO ANALYZE MANY MODELS.}
% In our review of the literature featuring cholera modeling in Haiti, we found that only a handful of papers that utilized mechanistic models reported more than a single model variation. (TODO: Cite).
% We acknowledged the possibility that a study reporting merely a single model may have explored others and only reported the one used in the final analysis.
% However, divulging details about the models that were considered, but not used, may greatly benefit the broader scientific community.
% These details may guide future researchers to avoid model variations that have previously demonstrated to perform poorly on incidence data.
% Furthermore, it can enhance confidence in the reported model's ability to describe the system under investigation.
% This confidence stems from the demonstrated evidence that the chosen model offers a better description of the observed data compared to other reasonable model alternatives.}

\new{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric like AIC.
Nested model variations are particularly useful as they enable formal statistical testing of the nested features.
The cholera case study demonstrated the advantage of such a model variation in all three models.
For instance, we tested the inclusion of a trend in transmission rate over time in Model~1; in Model~2, we considered adding an additional phase parameter, which enables a shift in the seasonal peaks of cholera infections; in Model~3, we included additional parameters such as those associated with Hurricane Matthew.
In all these cases, we found strong statistical evidence in favor of these model variations over their simpler counterparts.}

\new{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model.
Mathematical models necessarily simplify complex dynamical systems in order to make inference possible.
Decision-making about which features of a system need to be incorporated in a model and which should be excluded can significantly influence the model-based conclusions.
However, these decisions can be tested using a likelihood-based framework.
Standard model decisions cover deterministic versus stochastic modeling and the decision to include or exclude a meta-population structure.
Model benchmarks can assist in determining the level of complexity to be used in the model, enabling researchers to gauge a model's relative performance against a better-understood statistical model.}

\new{The three models in this case study offer a unique combination of decisions of model complexity.
Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm.
One of only a few examples of fitting a model with this level of complexity, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}

>>>>

\old{Recent information about a dynamic system should be more relevant for a forecast than older information. This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance. Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations [34]. That may explain why Lee et al.~[1] chose to obtain forecasts by simulating the calibrated models forward from initial conditions. We obtain forecasts by simulating future values using latent states that are consistent with the most recent observations; this is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$. The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~[1]: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.}

\new{\editForecastOne}

\new{\editForecastTwo}

@@

Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
\new{\editModVacc}
Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination. 
The results of these projections are summarized in Figs. \ref{fig:elimProbs}.

@@

\new{\editDiscussionOne}

\new{\editDiscussionTwo}

\new{\editDiscussionThree}

\new{\editDiscussionFour}

\new{\editDiscussionFive}
